Synthesizing Symmetric Lenses

### 2019-02-05 papers/SymmetricLenses/main.tex
update to feedback

\newif\ifdraft\drafttrue  % set true to show comments
% \newif\ifdraft\draftfalse  % set true to show comments
\newif\ifanon\anonfalse    % set true to suppress names, etc.
\newif\ifappendices\appendicestrue

@@

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to both input and output data to be transferred
in both directions.
Since their introduction by~\citet{Focal2005-long}, lenses have been extensively
studied, extended, and applied.  Recent work~\cite{optician} has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.  \bcp{Why is the next line not indented?}

@@

\noindent
In today's data-rich world, similar information is often
stored in different places and in different formats. For instance, electronic
calendars come in iCalendar, vCalendar, and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax
XML (used by the US government for e-filing) or TXF (used by
TurboTax)~\cite{tax-formats}; and sewing machines can use embroidery files
in formats with names like PES, DST, and several
others~\cite{embroidery-formats}.
%
\bcp{I'm not satisfied with the embroidery example: It doesn't obviously
  correspond to a bidirectional situation (the sewing machine does not need
  to send back the embroidery file to anybody).  The tax example could be
  criticized on the same grounds.}
%
\bcp{Also, it would be nice to be able to give a quick example of a truly
  symmetric case---do any of these examples work for that?}\afm{h-event and
  vCalendar are truly symmetric. Even the fields related to location alone
  are truly symmetric (though it is subtle to see this).}

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate building
principled data transformations.

\textit{Synthesizing} lenses 
can make the task even easier.  Indeed, lens languages are excellent
targets for search-based synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
limit the space of possible programs. 
%
In particular, the Optician~\cite{optician} algorithm\bcp{there's a
  potential for confusion between ``Optician'' and ``Boomerang''---are they
  distinct systems?  Is one part of the other?  Is one a system and one just
  an algorithm?  It's not very clear.} has demonstrated that
it is feasible to synthesize quite complex examples, albeit only for a
restricted class of lenses---so-called bijective lenses. Specifically, given 
source and target formats plus a small number of
examples of corresponding data, Optician's synthesis algorithm is guaranteed
to find a bijective lens matching the examples, if one exists. Indeed, 
it can often find an appropriate lens in less than a second with
no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

@@

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? In particular, what about \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure $S$ to a target $T$ can throw away information that must
then be restored when transferring from $T$ to $S$), and even more flexibile
transformations that allow each side to retain information not present in
the other?

\bcp{We use $S$ and $T$ here but $X$ and $Y$ below.}

@@

The first is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: the source $S$, the target $T$ and a complement
$C$ that contains the information not
present in either $S$ or $T$. Logistically, these complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses, which we call \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $S$ or $T$ when making a round trip. They only need the most
recent instance.

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly the symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  To check that they are expressive
enough in practice, we add simple symmetric lenses to the Boomerang
language~\cite{Boomerang}\bcp{this seems to assume that readers already know what
  Boomerang is} and apply them to a range of real-world
applications, showing in the process that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

@@

We consider a ``likely'' lens to be one that propagates ``a lot'' of
information from source to 
target and vice versa. Conversely, an unlikely lens is one that requires a
large amount of additional information to recover the target given a
source.\bcp{I'm nervous about the way we're talking about source and target
  here.  For one thing, even the words ``source'' and ``target'' connote a
  directionality that is misleading here.  And even ignoring that issue, a
  lot of the discussion (like here) sounds as though we are talking about
  unidirectional transformations.}  
This categorization follows from the intuition
that users will typically want to synchronise formats with lots
of shared information\bcp{really?  It's hard to be certain without lots of
  practical experience, but a priori it seems equally reasonable for users
  to want to pull little bits of shared information out of huge structures
  and keep them in sync.  But maybe we can weaken the claim and just say
  that we have an intuition that this will at least be {\em one} large class
  of application scenarios, and this is the one we focus on}, and so lenses
that propagate more information are preferable to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the shared information content of
two formats and our measure of likelihood.

@@

With simple symmetric lenses and our SRE-based likelihood measure in
hand, we develop a new algorithm for synthesizing likely lenses.
At its core, the algorithm performs a type-directed search between the 
descriptions of the source and target formats, measuring success using
the likelihood measure.
The interesting complications arise from the need to deal with
regular-expression equivalences.  There are infinitely many regular
expressions equivalent to a given one, and the lens returned by a
type-directed search will in general depend on which of the possible
representations are chosen for its source and target formats.  Moreover,
there may be {\em no} lens connecting the 
source and target formats selected\bcp{until we get past the colon, it is
  not clear that we are talking here about formats {\em not} modulo
  equivalence.} by the user: in general, the synthesis 
algorithm has to search to find equivalent ones for which a suitable lens
exists.  
To tame this complexity, we divide the synthesis algorithm into two
communicating search procedures (Figure~\ref{fig:high-level-algorithm}),
following~\citet{optician}. 
The first, \Expand, uses rewriting rules to iteratively
propose new pairs of stochastic regular expressions equivalent to the
original pair.
The second, \GreedySynth, takes a pair of SREs as input
and uses a greedy, type-directed algorithm to find a simple symmetric
lens between them, returning the lens and its likelihood score
to \Expand.  The whole synthesis algorithm terminates when a comparison of the
likilihood score of the best lens found so far with 
the complexity\bcp{what does it mean to compare likelihood with complexity?}
of newly generated SREs  
suggests that further searching is unlikely to be fruitful.

@@

We have added this synthesis procedure to the Boomerang system and evaluated
its effectiveness on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).
The ``programmer''\bcp{user?} can control the search process via examples and annotations
that informs the synthesis algorithm as to whether certain bits of information
are irrelevant, or must maintained.

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We
present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give
an algorithm for synthesizing lenses built from these combinators, using a 
novel application of stochastic regular expressions to guide the search
process (\S\ref{sec:synthesis})\bcp{``and prove it correct''!  (right?)}.  
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices\bcp{The appendices are pretty
  hideous and hard to navigate!!}
\ifappendices\else included with the supplemental material\bcp{Right?}\fi.

@@

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\vskip 1.5ex
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \tag{1}
 \mbox{$\CreateR \OfType X \rightarrow Y$}\\
 \tag{2}
 \mbox{$\CreateL \OfType Y \rightarrow X$}\\
 \tag{3}
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$}\\
 \tag{4}
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$}
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\vskip 1ex
\bcp{Are we going to use the numbers next to the laws??  Seems like not, so
  why have them at all?}
\noindent
The two \KW{create} functions are used to propagate shared information and
fill in default values when introducing new data (\EG, the ``unk'' salary
entry is added\bcp{is added by what?  A create function, I guess, but the
  connection is not clear} alongside the name to the management file when HR inserts a
new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the
other.  The round-tripping laws guarantee that pushing an ``un-edited''
value from one side (the result of a put or create) back through a lens in
the other direction will get back to exactly where we began.  \bcp{That
  paragraph seems likely to be quite opaque to any reader that is not
  already familiar with (symmetric) lenses.  How about a concrete example?}

This definition differs from classical symmetric lenses in that it does not
involve a \emph{complement}.  We give a detailed comparison in
\S\ref{sec:relationship}.

@@

In principle, programmers can define simple symmetric lenses just by
writing four arbitrary functions by hand and showing they satisfy the round-tripping
laws.  But a better idea is to streamline the process by providing a set of
combinators---a domain-specific 
language---for building law-abiding complex lenses from simpler ones.  Such
bidirectional languages
have been given for building lenses over many different kinds of data,
including database relations~\cite{BohannonPierceVaughan} and algebraic data
types~\cite{symmetric-lenses}\bcp{and what else? graphs, etc.?}.  We focus
here on combinators for defining lenses between string data formats
described by regular expressions.

@@

In contrast, the \Disconnect lens does not propagate any edits at all
from one format to the other.  It takes four arguments: two
regular expressions and two strings. The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). In our extended
example, the \lstinline{salary} field is only present in management files,
so we use the disconnect lens to ensure salary edits do not cause updates to the
HR file.
The defaults are strings that indicate the information is missing.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, the composite lens \lstinline{name_lens} in the running
example
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
will transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.
\bcp{We don't actually give a nontrivial example of how disconnect works.}

@@

\begin{figure}
\begin{lstlisting}
let name_lens = id(name) . id(" ") . id(name) . del(": ") . ins(",") in
let employee_lens = name_lens . disconnect(salary,"","unk","") 
                                                    . disconnect("",company,"","UNK") in
let employees_lens = ins("\n") . employee_lens . iterate(id("\n") . employee_lens) in
ins(header) . employees_lens : emp_salaries $\Leftrightarrow$ emp_insurance
\end{lstlisting}
%\ksf{I don't think this is necesssary.}
%let emp_lens = or(id(""),employees_lens) in
  \caption{A lens that synchronizes management and HR employee files\bcp{I'm
    pretty sure there was an error in name\_lens: the thing between the two
    names should be id(" ") (it was ins('','')).  Crossing my fingers that I'm
    making a correct fix...}}
  \label{fig:example_lens}
\end{figure}

@@

We provide typing rules for the simple symmetric combinators, most of
them syntax directed. However, regular expressions have a number of
equivalences on them, and sometimes we must convert one regular expression type
to another, equivalent, regular expression type. \bcp{The next sentence is
  dense.  Just show the rule?}Our typing rules include type
equivalence, so if a lens has type $\BRegex \Leftrightarrow \BRegexAlt$, then it
is also has type $\BRegex' \Leftrightarrow \BRegexAlt'$, where $\BRegex$ is
equivalent to $\BRegex'$, and $\BRegexAlt$ is equivalent to
$\BRegexAlt'$. \bcp{We're not talking about synthesis yet, so this next
  sentence seems off-track.  Drop it?}The
flexibility this type equivalence rule provides is useful when a different
representation of a regular expression would facilitate synthesis.

@@

That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.
\bcp{Suggest deleting this sentence:}The algorithm
uses such examples to find suitable defaults for \Disconnect lenses and to
figure out how to match data that could be synchronised in multiple ways.

@@

To obtain SREs from the plain regular expressions that users write, we use a
heuristic that attempts to assign probability annotations that give each
string in the language equal probablity. Our synthesis algorithm could
equally well take as input a\bcp{pair of} stochastic regular expression rather than a
plain one, permitting users to specify a data-specific probability
distribution (either manually calculated or inferred from some dataset of
interest). However, we have found that our heuristic works well in practice,
so we have not explored these options.

>>>>

\newif\ifdraft\drafttrue  % set true to show comments
% \newif\ifdraft\draftfalse  % set true to show comments
\newif\ifanon\anonfalse    % set true to suppress names, etc.
\newif\ifappendices\appendicesfalse

@@

{\em Lenses} are programs that can be run both ``front to back'' and ``back
to front,'' allowing updates to both input and output data to be transferred
in both directions.
Since their introduction by~\citet{Focal2005-long}, lenses have been extensively
studied, extended, and applied.  Recent work~\cite{optician} has also demonstrated how
techniques from {\em type-directed program synthesis} can be used to
efficiently synthesize a simple class of lenses---so-called {\em
  bijective lenses} over string data---given a pair of types (regular
expressions) and a small number of examples.  \bcp{Why is the next line not
  indented?} \afm{I have truly no idea, deferring until I get a stackoverflow answer.}

@@

\noindent
In today's data-rich world, similar information is often
stored in different places and in different formats. For instance, electronic
calendars come in iCalendar, vCalendar, and h-event
formats~\cite{calendar-formats}; US taxation data can be transmitted via Tax
XML (used by the US government for e-filing) or TXF (used by
TurboTax)~\cite{tax-formats}; and sewing machines can use embroidery files
in formats with names like PES, DST, and several
others~\cite{embroidery-formats}.
%
\bcp{I'm not satisfied with the embroidery example: It doesn't obviously
  correspond to a bidirectional situation (the sewing machine does not need
  to send back the embroidery file to anybody).  The tax example could be
  criticized on the same grounds.}
\afm{I still think that there are situations where the embroidery file makes
  sense (I'm sharing my design with a friend, who then tweaks it and sends it
  back. There may also need to file-specific information, where one contains
  parameters like how fast to embroider and some doesn't, to be honest I don't
  know much about embroidery though. Thinking about other examples, however.)}

One convenient way to maintain consistency between such varied data formats
is to use a \emph{lens}~\cite{Focal2005-long}---a bidirectional program that
can transform updates to data represented in a format $S$ to another format
$T$ and vice versa, providing ``round tripping'' guarantees that help ensure
information is not lost or corrupted as it is transformed back and forth
between different representations.
Domain-specific languages for writing lenses can facilitate building
principled data transformations.  For example, Boomerang~cite{boomerang} is a
language for expressing bidirectional string transformations.

\textit{Synthesizing} lenses can make the task even easier. Indeed, lens
languages, like Boomerang, are excellent targets for search-based synthesis, since their
specialized, sub-Turing-complete syntax and expressive type systems drastically
limit the space of possible programs.
%
In particular, the recent Optician~\cite{optician} enhancement to Boomerang has
demonstrated that it is feasible to synthesize quite complex examples, albeit
only for a restricted class of lenses---so-called bijective lenses.
Specifically, given source and target formats plus a small number of examples of
corresponding data, Optician's synthesis algorithm is guaranteed to find a
bijective lens matching the examples, if one exists. Indeed, it can often find
an appropriate lens in less than a second with no examples at all.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis can be so effective, even relative to
successful synthesis tools in other domains, is that there are typically not
very many bijections between a given pair of data formats. 
If the synthesis algorithm can
find any bijection at all, it is fairly likely to be the intended one.

@@

Can Optician's basic synthesis procedure be extended to a richer class of
lenses? In particular, what about \emph{symmetric
  lenses}~\cite{symmetric-lenses}---a much larger class that includes
bijective lenses, ``asymmetric'' lenses (where the transformation from a
source structure to a target can throw away information that must
then be restored when transferring from target to source), and even more flexibile
transformations that allow each side to retain information not present in
the other?

@@

The first is pragmatic. ``Classical'' symmetric lenses as presented
by ~\citet{symmetric-lenses} operate
over three structures: the ``left'' structure $X$, the ``right'' $Y$ and a complement
$C$ that contains the information not
present in either $X$ or $Y$. Logistically, these complements must be stored
and managed somehow.  More fundamentially, complements complicate
synthesis {\em specifications}---instead of just giving single examples of
source and target 
pairings, users would have to give longer ``interaction sequences'' to 
show how a lens should behave. To avoid these complexities, we define a
restricted variant 
of symmetric lenses, which we call \emph{simple symmetric lenses}. 
%only the usual source and target.
%
%, which we define in Section~\ref{sec:relationship}.
Intuitively, simple symmetric lenses are the complete set of symmetric
lenses that do not require external ``memory'' to recover data from past
instances of $X$ or $Y$ when making a round trip. They only need the most
recent instance.

To characterize their expressive power formally, we show that simple
symmetric lenses are exactly the symmetric lenses that satisfy an intuitive
property called \emph{forgetfulness}.  To check that they are expressive
enough in practice, we add simple symmetric lenses to the Boomerang
language~\cite{Boomerang} and apply them to a range of real-world
applications, showing in the process that simple symmetric lenses integrate
smoothly with other advanced lens features provided by Boomerang, including
quotient lenses~\cite{quotientlenses,maina+:quotient-synthesis} and matching
lenses~\cite{matchinglenses}.

@@

We consider a ``likely'' lens to be one that propagates ``a lot'' of information
from the left data format to the right and vice versa. Conversely, an unlikely
lens is one that requires a large amount of additional information to recover
either of the formats given the other. For many application scenarios, users
will want to synchronise formats with lots of shared information (and lenses
with little shared information are typically easier to write). Our tool is built
primarily for these complex scnearios, and so lenses that propagate more
information are preferable in our context to ones that propagate less.
%
This intuition is formalized using \emph{stochastic regular expressions}
(SREs)~\cite{stoch-rnn,stoch-def}, which simultaneously define a set of
strings and a probability distribution over those strings.  Using this
probability distribution, we can calculate the shared information content of
two formats and our measure of likelihood.

@@

With simple symmetric lenses and our SRE-based likelihood measure in hand, we
develop a new algorithm for synthesizing likely lenses. At its core, the
algorithm performs a type-directed search between the descriptions of the source
and target formats, measuring success using the likelihood measure. The
interesting complications arise from the need to deal with regular-expression
equivalences. There are infinitely many regular expressions equivalent to a
given one, and the lens returned by a type-directed search will in general
depend on which of the possible representations are chosen for its source and
target formats. Moreover, if Boomerang's type-equivalence rule is disallowed,
there may be {\em no} well-typed lens connecting the source and target formats
selected by the user: in general, the synthesis algorithm has to search through
equivalent regular expression types for which a suitable lens exists. To tame
this complexity, we divide the synthesis algorithm into two communicating search
procedures (Figure~\ref{fig:high-level-algorithm}), following~\citet{optician}.
The first, \Expand, uses rewriting rules to iteratively propose new pairs of
stochastic regular expressions equivalent to the original pair. The second,
\GreedySynth, takes a pair of SREs as input and uses a greedy, type-directed
algorithm to find a simple symmetric lens between them, returning the lens and
its likelihood score to \Expand. The whole synthesis algorithm heuristically
terminates when the most likely lens is no longer unlikely enough to justify
the time cost of searching through newly generated SREs.

@@

We have added this synthesis procedure to the Boomerang system and evaluated
its effectiveness on a benchmark suite of 48 problems, including data-cleaning
tasks taken from Flash Fill~\cite{gulwani-popl-2014}, view-maintenance tasks
taken from Augeas~\cite{augeas2}, and ad-hoc synchronization tasks from a
variety of other sources (mostly from the bidirectional programming literature).
The user can control the search process via examples and annotations
that informs the synthesis algorithm as to whether certain bits of information
are irrelevant, or must maintained.

In summary, our contributions are:
\begin{itemize}
%
\item We define {\em simple symmetric lenses}, a subclass of symmetric
lenses that do not maintain any hidden state (\S\ref{sec:overview}).  
%
We present a collection of combinators for building simple symmetric lenses and
describe how to calculate their likelihoods (\S\ref{sec:ssl}).
\item We give an algorithm for synthesizing lenses built from these combinators,
  using a novel application of stochastic regular expressions to guide the
  search process (\S\ref{sec:synthesis}) and prove that our metrics have deep
  semantic meaning.
%
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with other lens extensions such as
  quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}. We evaluate our implementation, the effects of
  various optimizations, and our inclusion of relevance annotations on a set of
  48 lens synthesis benchmarks drawn from data cleaning, view maintenance, and
  file synchronization tasks. The system can synthesize simple symmetric lenses
  for all of the benchmarks in under 30 seconds (\S\ref{sec:evaluation}).
%
\item We show that simple symmetric lenses fall between asymmetric
lenses~\cite{Focal2005-long} and full symmetric
lenses~\cite{symmetric-lenses} in expressiveness.  Indeed, the
class of simple lenses can be characterized semantically as the subset of
full symmetric lenses for which a simple ``forgetfulness'' property holds
(\S\ref{sec:related}).
\end{itemize}
Along the
way, we show that {\em star-semiring equivalences} on regular expressions
remain valid when extended to stochastic regular expressions, preserving
probability distributions as well as languages (\S\ref{sec:sre}).
We close with related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).
Elided proofs can be found in the appendices\bcp{The appendices are pretty
  hideous and hard to navigate!!}
\ifappendices\else included with the supplemental material\fi.

@@

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\vskip 1.5ex
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \mbox{$\CreateR \OfType X \rightarrow Y$} \nonumber \\
 \mbox{$\CreateL \OfType Y \rightarrow X$} \nonumber \\
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$} \nonumber \\
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$} \nonumber
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\vskip 1ex
\noindent
The two \KW{create} functions are used to propagate shared information and fill
in default values when introducing new data (\EG, on \KW{create} the ``unk''
salary entry is added alongside the name to the management file when HR inserts
a new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the other.
The notation $\Lens.\PutR$ extracts the \PutR function from simple symmetric
lens $\Lens$; similarly for \PutL. These four functions can be used to keep the
common information between two file types in synch. For example, if a new file
of the left-hand format is created, the \CreateR function will a new file in the
right-hand format. If the right-hand format is edited, the \PutL function can
update the left-hand format with the changed information.

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider,
\lstinline{emp_lens}, a lens between
\lstinline{emp_salaries} and
\lstinline{emp_insurance}. Applying
\lstinline{emp_lens}.\CreateR to a salary file creates a insurance file with the
same employee names in the file, with \lstinline{"UNK"} for each insurance
company. The round-tripping laws guarantee that using \lstinline{emp_lens}.\PutL
to put the generated insurance file into the original salary file will return
the original salary file, unmodified. If \KW{put}ting a string $s$ into a string
$t$ results in the original string $s$, then $s$ and $t$ are considered
\emph{synchronized}.

This simple symmetric lens definition differs from classical symmetric lenses in
that it does not involve a \emph{complement}. We give a detailed comparison in
\S\ref{sec:relationship}.

@@

In principle, programmers can define simple symmetric lenses just by writing
four arbitrary functions by hand and showing they satisfy the round-tripping
laws. But a better idea is to streamline the process by providing a set of
combinators---a domain-specific language---for building law-abiding complex
lenses from simpler ones. Such bidirectional languages have been given for
building lenses over many different kinds of data, including database
relations~\cite{BohannonPierceVaughan} algebraic data
types~\cite{symmetric-lenses}, graphs~\cite{tgg}, and more.

We focus here on combinators for defining lenses between string data formats
described by regular expressions.

@@

In contrast, the \Disconnect lens does not propagate any edits at all
from one format to the other.  It takes four arguments: two
regular expressions and two strings. The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). In our extended
example, the \lstinline{salary} field is only present in management files,
so we use the disconnect lens to ensure salary edits do not cause updates to the
HR file.
For example, with the above lens, $\PutL \App \lstinline{""} \App 60000$ will return 60000,
and \PutR will always return \lstinline{""}.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, the composite lens \lstinline{name_lens} in the running
example
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
will transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.

@@

\begin{figure}
\begin{lstlisting}
let name_lens = id(name) . id(" ") . id(name) . del(": ") . ins(",") in
let employee_lens = name_lens . disconnect(salary,"","unk","") 
                                                    . disconnect("",company,"","UNK") in
let employees_lens = ins("\n") . employee_lens . iterate(id("\n") . employee_lens) in
ins(header) . employees_lens : emp_salaries $\Leftrightarrow$ emp_insurance
\end{lstlisting}
%\ksf{I don't think this is necesssary.}
%let emp_lens = or(id(""),employees_lens) in
  \caption{A lens that synchronizes management and HR employee files}
  \label{fig:example_lens}
\end{figure}

@@

We provide typing rules for the simple symmetric combinators, most of
them syntax directed. However, regular expressions have a number of
equivalences on them, and sometimes we must convert one regular expression type
to another, equivalent, regular expression type.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.

@@

To obtain SREs from the plain regular expressions that users write, we use a
heuristic that attempts to assign probability annotations that give each
string in the language equal probablity. Our synthesis algorithm could
equally well take as input a pair of stochastic regular expression rather than a
plain one, permitting users to specify a data-specific probability
distribution (either manually calculated or inferred from some dataset of
interest). However, we have found that our heuristic works well in practice,
so we have not explored these options.

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that we can use types to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions. The fact that each regular expression is semantically equivalent to
infinitely many others complicates the problem.
Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searches. The first search,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both the semantics and the
probability distributions of those stochastic regular expressions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.
